Plant trait cleaning and combining

### 2024-05-30 code/L1/L1_P4_TropicalAndes_plant_traits.Rmd
add new packages, update input files, modify functions to include record count and variance, include summary figures to see trait coverage before imputation

# Load required packages
```{r}
library(dplyr)
library(tidyr)
```

# Set file paths
```{r}
data_path_L0<-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L0')
data_path_L1 <-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
output_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
```

# Read in subset fruiting species data
```{r}
plant_occ <- read.csv(file.path(data_path_L1, "TropicalAndes_GBIF_plant_occ_harmonized.csv"))
TRY_traits <- read.csv(file.path(data_path_L1,"TropicalAndes_TRY_plant_traits_harmonized.csv"))
BIEN_traits <- read.csv(file.path(data_path_L1,"TropicalAndes_BIEN_plant_traits_harmonized.csv"))
GIFT_traits <- read.csv(file.path(data_path_L1,"TropicalAndes_GIFT_plant_traits_harmonized.csv"))
```

# Look at trait numbers
```{r}
traits %>%
  count(TraitName)
```

```{r}
traits %>%
  distinct(Accepted_species, TraitName) %>%
  count(TraitName)
```

Final trait list:
- dispersal syndrome
- Fruit mass
- Fruit length
- Fruit type
- Plant growth form
- Plant height
- Plant Lifespan
- Seed mass
- Seed length
- Seed width

# Standardize trait measurements
## Create functions
```{r}
# Numeric data
numeric_data_average <- function(data, unit, traitname) {
# remove rows if TraitValue is non-numeric
data <- data %>%
  filter(!is.na(as.numeric(TraitValue))) %>%
  mutate(TraitValue = as.numeric(TraitValue))
# remove duplicate rows
data <- data %>%
  distinct(.keep_all = TRUE)
# calculate average value per trait per species
average_data <- aggregate(TraitValue ~ Accepted_species, data, FUN = mean, na.rm = TRUE)
# Remove rows with missing or empty species names
average_data <- average_data[!is.na(average_data$Accepted_species) & average_data$Accepted_species !="", ]
# Add unit & trait column
average_data$Unit <- unit
average_data$TraitName <- traitname

return(average_data)
}
```

```{r}
# categorical data
factor_data_merge <- function(data, traitname) {
# remove duplicate rows
data <- data %>%
  distinct(.keep_all = TRUE)
# remove rows with missing or empty species names
data <- data[!is.na(data$Accepted_species) & data$Accepted_species != "", ]
# Subset data to just species and TraitValue
data <- data[,c("Accepted_species", "TraitValue")]
# Find the most common trait value for each species
  most_common_values <- aggregate(. ~ Accepted_species, data, function(x) {
    # Find the mode (most common value) of the trait values
    mode_value <- names(sort(table(x), decreasing = TRUE))[1]
    return(mode_value)
  })
# Merge most common trait values back to the original data
merged_data <- merge(data, most_common_values, by = "Accepted_species", all.x = TRUE)

# Rename the merged column
colnames(merged_data)[ncol(merged_data)] <- "TraitValue"

# Remove column TraitValue.x
merged_data <- merged_data[, !colnames(merged_data) %in% "TraitValue.x", drop = FALSE]

# remove duplicate rows
merged_data <- merged_data %>%
  distinct(.keep_all = TRUE)

# Add unit & trait column
merged_data$TraitName <- traitname

  return(merged_data)
}
```

## Dispersal syndrome
```{r}
traits_dispersal <- traits %>%
  filter(grepl('Dispersal syndrome|whole plant dispersal syndrome|dispersal_syndrome_1|dispersal_syndrome_2', TraitName))
traits_dispersal$OrigTraitValue <- traits_dispersal$TraitValue
nrow(traits_dispersal)
length(unique(traits_dispersal$Accepted_species))
```

@@

## Fruit dry mass
```{r}
traits_fruitMass <- traits %>%
  filter(grepl('Fruit dry mass', TraitName))
traits_fruitMass$OrigTraitValue <- traits_fruitMass$TraitValue
nrow(traits_fruitMass)
length(unique(traits_fruitMass$Accepted_species))
```

@@

## Fruit length
```{r}
traits_fruitLength <- traits %>%
  filter(grepl('Fruit length|maximum fruit length|minimum fruit length|fruit_length_min|fruit_length_max|fruit_length_mean', TraitName))
traits_fruitLength$OrigTraitValue <- traits_fruitLength$TraitValue

@@

## Plant height
```{r}
traits_plantHeight <- traits %>%
  filter(grepl('Plant height generative|Plant height vegetative|maximum whole plant height|minimum whole plant height|plant_height|plant_height_max|plant_height_min|whole plant height', TraitName))
traits_plantHeight$OrigTraitValue <- traits_plantHeight$TraitValue

@@

## Growth form
```{r}
traits_growthForm <- traits %>%
  filter(grepl('Plant growth form|growth_form|whole plant growth form'
, TraitName))
traits_growthForm$OrigTraitValue <- traits_growthForm$TraitValue

@@

```{r}
averagetraits_growthForm <- factor_data_merge(traits_growthForm, "Growth Form")
```

# Combine all traits
```{r}
# add unit column to categorical traits
cat_traits <- rbind(averagetraits_dispersal, averagetraits_fruitType, averagetraits_growthForm)
cat_traits$Unit <- ""
clean_traits <- rbind(averagetraits_fruitLength, averagetraits_fruitMass, averagetraits_plantHeight, averagetraits_plantLifespan, averagetraits_seedLength, averagetraits_seedMass, averagetraits_seedWidth, cat_traits)
nrow(clean_traits)
```

@@

# Summary
```{r}
print("Number of records: ")
nrow(clean_traits)
print("Number of species: ")
length(unique(clean_traits$Accepted_species))
```

# Get one trait value per species
```{r}
#remove unit column from dataframe
clean_traits <- clean_traits[, !colnames(clean_traits) %in% "Unit", drop = FALSE]
```

```{r}
# separate numeric and non-numeric traits
numeric_traits <- clean_traits[clean_traits$TraitName %in% c("Fruit Length", "Fruit Mass", "Plant Height", "Plant Lifespan", "Seed Length", "Seed Width", "Seed Mass"), ]
cat_traits <- clean_traits[clean_traits$TraitName %in% c("Dispersal Syndrome", "Fruit Type", "Growth Form"), ]

numeric_traits$TraitValue <- as.numeric(numeric_traits$TraitValue)
```

```{r}
# numeric traits
## Group by species and TraitName, and calculate the mean TraitValue for each combination
numeric_traits_combined <- numeric_traits %>%
  group_by(species, TraitName) %>%
  summarise(TraitValue = mean(TraitValue, na.rm = TRUE)) %>%
  ungroup()

# non-numeric traits
cat_traits_combined <- cat_traits %>%
  group_by(species) %>%
  summarise_all(.funs = list(mode_trait = function(x) {
    mode_value <- names(sort(table(x), decreasing = TRUE))[1]  # Calculate mode
    return(mode_value)
  }))
```

```{r}
#rename columns to original values
colnames(cat_traits_combined)[which(colnames(cat_traits_combined) == "TraitValue_mode_trait")] <- "TraitValue"
colnames(cat_traits_combined)[which(colnames(cat_traits_combined) == "TraitName_mode_trait")] <- "TraitName"

# Remove rows with "?" or numbers in the TraitValue column
cleaned_cat_traits <- cat_traits_combined %>%
  filter(!grepl("\\?", TraitValue) & !grepl("\\d", TraitValue))
```

```{r}
#combine numeric and non-numeric traits
combined_clean_traits <- rbind(cat_traits_combined, numeric_traits_combined)
```

```{r}
combined_clean_traits %>%
  count(TraitName)
```

# Remove traits with less than 10 records
```{r}
combined_clean_traits <- combined_clean_traits[!(combined_clean_traits$TraitName %in% c("Fruit Mass", "Seed Mass")), ]
```

# Transform plant dataframe from long to wide
```{r}
wide_traits <- spread(combined_clean_traits, key = TraitName, value = TraitValue)

#remove X column from dataframe
wide_plant_traits <- wide_traits[, !colnames(wide_traits) %in% "X", drop = FALSE]
```

```{r}
# Assign data types
wide_plant_traits$`Fruit Length` <- as.numeric(wide_plant_traits$`Fruit Length`)
wide_plant_traits$`Plant Height` <- as.numeric(wide_plant_traits$`Plant Height`)
wide_plant_traits$`Dispersal Syndrome` <- as.factor(wide_plant_traits$`Dispersal Syndrome`)
wide_plant_traits$`Growth Form` <- as.factor(wide_plant_traits$`Growth Form`)
wide_plant_traits$`Fruit Type` <- as.factor(wide_plant_traits$`Fruit Type`)
```

# Summary
```{r}
print("Number of species: ")
length(unique(wide_plant_traits$species))
```

# Write data to csv
```{r}
write.csv(wide_plant_traits, file.path(output_path,"TropicalAndes_all_plant_traits_harmonized.csv"))
write.csv(clean_traits, file.path(output_path,"TropicalAndes_all_plant_traits_harmonized_with_units.csv"))
```

>>>>

# Load required packages
```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(funbiogeo)
library(visdat)
```

# Set file paths
```{r}
data_path_L0<-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L0')
data_path_L1 <-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
output_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
figure_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/figures')
```

# Read in subset fruiting species data
```{r}
plant_traits <- read.csv(file.path(data_path_L1, "TropicalAndes_all_plant_traits_harmonized_subset.csv"))
```

# Look at trait numbers
```{r}
plant_traits %>%
  count(TraitName)
```

```{r}
plant_traits %>%
  distinct(Accepted_species, TraitName) %>%
  count(TraitName)
```

Final trait list:
- dispersal syndrome
- Fruit mass
- Fruit length
- Plant growth form
- Plant height
- Plant Lifespan
- Seed mass
- Seed length
- Seed width

# Standardize trait measurements
## Create functions
```{r}
numeric_data_average <- function(data, unit, traitname) {
  # Remove rows if TraitValue is non-numeric and convert to numeric
  data <- data %>%
    filter(!is.na(as.numeric(TraitValue))) %>%
    mutate(TraitValue = as.numeric(TraitValue))

  # Remove duplicate rows
  data <- data %>%
    distinct(.keep_all = TRUE)

  # Remove rows with missing or empty species names
  data <- data[!is.na(data$Accepted_species) & data$Accepted_species != "", ]

  # Calculate average value, count, and variance per trait per species
  summary_data <- data %>%
    group_by(Accepted_species) %>%
    summarise(
      TraitValue_mean = mean(TraitValue, na.rm = TRUE),
      records_used = n(),
      variance = var(TraitValue, na.rm = TRUE)
    )

  # Add trait name column
  summary_data$TraitName <- traitname

  # Rename the mean column to TraitValue
  summary_data <- summary_data %>%
    rename(TraitValue = TraitValue_mean)

  return(summary_data)
}

```{r}
factor_data_merge <- function(data, traitname) {

  # Remove rows with missing or empty species names
  data <- data[!is.na(data$Accepted_species) & data$Accepted_species != "", ]

  # Subset data to just species and TraitValue
  data <- data[, c("Accepted_species", "TraitValue")]

  # Calculate mode, count, and variance-like measure (using mode frequency)
  summary_data <- data %>%
    group_by(Accepted_species) %>%
    summarise(
      ModeValue = names(sort(table(TraitValue), decreasing = TRUE))[1],
      records_used = n(),
      mode_freq = max(table(TraitValue)),
      variance = 1 - (max(table(TraitValue)) / n())
    )

  # Add trait name column
  summary_data$TraitName <- traitname

  # Rename the mean column to TraitValue
  summary_data <- summary_data %>%
    rename(TraitValue = ModeValue)

  return(summary_data)
}
```

## Dispersal syndrome
```{r}
traits_dispersal <- plant_traits %>%
  filter(grepl('Dispersal syndrome|whole plant dispersal syndrome|dispersal_syndrome_1|dispersal_syndrome_2', TraitName))
traits_dispersal$OrigTraitValue <- traits_dispersal$TraitValue
nrow(traits_dispersal)
length(unique(traits_dispersal$Accepted_species))
```

@@

## Fruit dry mass
```{r}
traits_fruitMass <- plant_traits %>%
  filter(grepl('Fruit dry mass', TraitName))
traits_fruitMass$OrigTraitValue <- traits_fruitMass$TraitValue
nrow(traits_fruitMass)
length(unique(traits_fruitMass$Accepted_species))
```

@@

## Fruit length
```{r}
traits_fruitLength <- plant_traits %>%
  filter(grepl('Fruit length|maximum fruit length|minimum fruit length|fruit_length_min|fruit_length_max|fruit_length_mean', TraitName))
traits_fruitLength$OrigTraitValue <- traits_fruitLength$TraitValue

@@

## Plant height
```{r}
traits_plantHeight <- plant_traits %>%
  filter(grepl('Plant height generative|Plant height vegetative|maximum whole plant height|minimum whole plant height|plant_height|plant_height_max|plant_height_min|whole plant height', TraitName))
traits_plantHeight$OrigTraitValue <- traits_plantHeight$TraitValue

@@

## Growth form
Just using data from GIFT 
```{r}
traits_growthForm <- plant_traits %>%
  filter(TraitName == 'growth_form')
traits_growthForm$OrigTraitValue <- traits_growthForm$TraitValue

@@

```{r}
# standardize trait names with units
traits_growthForm <- traits_growthForm %>%
  mutate(TraitName = "GrowthForm")
```

```{r}
averagetraits_growthForm <- factor_data_merge(traits_growthForm, "GrowthForm")
```

```{r}
nrow(averagetraits_growthForm)
length(unique(averagetraits_growthForm$Accepted_species))
```

```{r}
summary(averagetraits_growthForm)
```

# Combine all traits
```{r}
trait_dfs <- list(averagetraits_fruitLength, averagetraits_fruitMass, averagetraits_plantHeight, averagetraits_plantLifespan, averagetraits_seedLength, averagetraits_seedMass, averagetraits_seedWidth, averagetraits_dispersal, averagetraits_growthForm)

# Function to combine trait dataframes only matching columns
combine_matching_columns <- function(df_list) {
  # Find the common columns across all data frames
  common_cols <- Reduce(intersect, lapply(df_list, colnames))

  # Ensure TraitValue column is of the same type across all data frames
  df_list <- lapply(df_list, function(df) {
    if ("TraitValue" %in% colnames(df)) {
      df <- df %>% mutate(TraitValue = as.character(TraitValue))
    }
    return(df)
  })

  # Filter each data frame to keep only the common columns
  df_list_filtered <- lapply(df_list, function(df) {
    df %>% select(all_of(common_cols))
  })

  # Combine the data frames using reduce and full_join
  combined_df <- reduce(df_list_filtered, full_join, by = common_cols)

  return(combined_df)
}
clean_traits <- combine_matching_columns(trait_dfs)
nrow(clean_traits)
```

@@

# Transform plant dataframe from long to wide
```{r}
wide_traits <- spread(clean_traits, key = TraitName, value = TraitValue)

#remove X column from dataframe
wide_plant_traits <- wide_traits[, !colnames(wide_traits) %in% c("X", "records_used", "variance"), drop = FALSE]
```

```{r}
# Assign data types
wide_plant_traits$`FruitLength_mm` <- as.numeric(wide_plant_traits$`FruitLength_mm`)
wide_plant_traits$`FruitMass_mg` <- as.numeric(wide_plant_traits$`FruitMass_mg`)
wide_plant_traits$`PlantHeight_m` <- as.numeric(wide_plant_traits$`PlantHeight_m`)
wide_plant_traits$`PlantLifespan_years` <- as.numeric(wide_plant_traits$`PlantLifespan_years`)
wide_plant_traits$`SeedLength_mm` <- as.numeric(wide_plant_traits$`SeedLength_mm`)
wide_plant_traits$`SeedMass_g` <- as.numeric(wide_plant_traits$`SeedMass_g`)
wide_plant_traits$`SeedWidth_mm` <- as.numeric(wide_plant_traits$`SeedWidth_mm`)
wide_plant_traits$`DispersalSyndrome` <- as.factor(wide_plant_traits$`DispersalSyndrome`)
wide_plant_traits$`GrowthForm` <- as.factor(wide_plant_traits$`GrowthForm`)
```

```{r}
# rename accepted_species column to species
wide_plant_traits <- wide_plant_traits %>%
  rename(species = Accepted_species)
```

# Summary
```{r}
cat("Number of records:", nrow(clean_traits), "\n")
cat("Number of species:", length(unique(clean_traits$Accepted_species)), "\n")
```

```{r}
fb_plot_species_traits_completeness(wide_plant_traits)
fb_plot_number_species_by_trait(wide_plant_traits)
fb_table_trait_summary(wide_plant_traits)
```

```{r}
vis_dat(wide_plant_traits)
vis_miss(wide_plant_traits)
```

# Write data to csv
```{r}
write.csv(wide_plant_traits, file.path(output_path,"TropicalAndes_all_plant_traits_standardized.csv"))
write.csv(clean_traits, file.path(output_path,"TropicalAndes_all_plant_traits_standardized_with_recordCount.csv"))
```
